---
title: "DE, ISC"
output: html_document
date: "2023-11-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Differential gene expression with DESeq2

```{r}
library(tidyverse)
library(cowplot)
library(Matrix.utils)
library(edgeR)
library(Matrix)
library(reshape2)
library(S4Vectors)
library(pheatmap)
library(apeglm)
library(png)
library(DESeq2)
library(RColorBrewer)
library(data.table)
library(SingleCellExperiment)
library(ggplot2)
library(ggrepel)
```

```{r}
# Bring in rab11_ctrl_combined object
rab11_ctrl_combined <- readRDS(file="/g/huber/users/lazoriak/scripts/rab11/rab11 analysis/rab11_ctrl.rds")

# Extract raw counts and metadata to create SingleCellExperiment object
counts <- rab11_ctrl_combined@assays$RNA@counts 

metadata <- rab11_ctrl_combined@meta.data

# Set up metadata as desired for aggregation and DE analysis
#before this step, make sure object is annotated
metadata$cluster_id <- factor(rab11_ctrl_combined$predicted.celltype)
metadata$cluster_id <- case_when (
  metadata$cluster_id %in% c("aEC", "pEC", "mEC", "LFC") ~ "EC",
  TRUE ~ as.character(metadata$cluster_id))

metadata <- metadata %>%
  mutate(
    cluster_id = case_when(
      cluster_id %in% c("unk2", "unk1", "unk3") ~ NA_character_,
      TRUE ~ cluster_id
    )
  )
metadata$cluster_id <- factor(metadata$cluster_id)
  
metadata$sample_id <- factor(rab11_ctrl_combined$orig.ident)

metadata$group_id <- factor(metadata$group_id)

# Create single cell experiment object
sce <- SingleCellExperiment(assays = list(counts = counts), 
                           colData = metadata)
```

```{r}
# Explore the raw counts for the dataset

## Check the assays present
assays(sce)

## Check the counts matrix
dim(counts(sce))
counts(sce)[1:6, 1:6]
```

```{r}
# Explore the cellular metadata for the dataset

dim(colData(sce))
head(colData(sce))
```

```{r}
#factor<-as.factor(sce$labels)
# Extract unique names of clusters (= levels of cluster_id factor variable)
#cluster_names <- levels(factor)
cluster_names <- levels(colData(sce)$cluster_id)
cluster_names

# Total number of clusters
length(cluster_names)
```

```{r}
# Extract unique names of samples (= levels of sample_id factor variable)

sample_names <- levels(colData(sce)$sample_id)
sample_names

# Total number of samples
length(sample_names)
```

```{r}
# Subset metadata to include only the variables you want to aggregate across (here, we want to aggregate by sample and by cluster)
groups <- colData(sce)[, c("cluster_id", "sample_id")]
head(groups)
```

```{r}
# Aggregate across cluster-sample groups
# transposing row/columns to have cell_ids as row names matching those of groups
aggr_counts <- aggregate.Matrix(t(counts(sce)), 
                                groupings = groups, fun = "sum") 

# Explore output matrix
class(aggr_counts)
dim(aggr_counts)
aggr_counts[1:6, 1:6]
```

```{r}
# Transpose aggregated matrix to have genes as rows and samples as columns
aggr_counts <- t(aggr_counts)
aggr_counts[1:6, 1:6]
```

```{r}
# Understanding tstrsplit()

## Exploring structure of function output (list)
tstrsplit(colnames(aggr_counts), "_") %>% str()

## Comparing the first 10 elements of our input and output strings
head(colnames(aggr_counts), n = 10)
head(tstrsplit(colnames(aggr_counts), "_")[[1]], n = 10)
```

```{r}
# Using which() to look up tstrsplit() output
ISC_idx <- which(tstrsplit(colnames(aggr_counts), "_")[[1]] == "ISC")
ISC_idx

colnames(aggr_counts)[ISC_idx]
aggr_counts[1:10, ISC_idx]
```

```{r}
# As a reminder, we stored our cell types in a vector called cluster_names
cluster_names


# Loop over all cell types to extract corresponding counts, and store information in a list

## Initiate empty list
counts_ls <- list()

for (i in 1:length(cluster_names)) {

  ## Extract indexes of columns in the global matrix that match a given cluster
  column_idx <- which(tstrsplit(colnames(aggr_counts), "_")[[1]] == cluster_names[i])
  
  ## Store corresponding sub-matrix as one element of a list
  counts_ls[[i]] <- aggr_counts[, column_idx]
  names(counts_ls)[i] <- cluster_names[i]

}

# Explore the different components of the list
str(counts_ls)
```

```{r}
# Reminder: explore structure of metadata
head(colData(sce))

# Extract sample-level variables
metadata <- colData(sce) %>% 
  as.data.frame() %>% 
  dplyr::select(sample_id, group_id)

dim(metadata)
head(metadata)

# Exclude duplicated rows
metadata <- metadata[!duplicated(metadata), ]

dim(metadata)
head(metadata)
```

```{r}
# Rename rows
rownames(metadata) <- metadata$sample_id
head(metadata)
```

```{r}
# Number of cells per sample and cluster
t <- table(colData(sce)$sample_id,
           colData(sce)$cluster_id)
t
```

```{r}
# Creating metadata list

## Initiate empty list
metadata_ls <- list()

for (i in 1:length(counts_ls)) {
  
    ## Initiate a data frame for cluster i with one row per sample (matching column names in the counts matrix)
    df <- data.frame(cluster_sample_id = colnames(counts_ls[[i]]))
    
    ## Use tstrsplit() to separate cluster (cell type) and sample IDs
    df$cluster_id <- tstrsplit(df$cluster_sample_id, "_")[[1]]
    df$sample_id  <- tstrsplit(df$cluster_sample_id, "_")[[2]]
    
    
    ## Retrieve cell count information for this cluster from global cell count table
    idx <- which(colnames(t) == unique(df$cluster_id))
    cell_counts <- t[, idx]
    
    ## Remove samples with zero cell contributing to the cluster
    cell_counts <- cell_counts[cell_counts > 0]
    
    ## Match order of cell_counts and sample_ids
    sample_order <- match(df$sample_id, names(cell_counts))
    cell_counts <- cell_counts[sample_order]
    
    ## Append cell_counts to data frame
    df$cell_count <- cell_counts
    
    
    ## Join data frame (capturing metadata specific to cluster) to generic metadata
    df <- plyr::join(df, metadata, 
                     by = intersect(names(df), names(metadata)))
    
    ## Update rownames of metadata to match colnames of count matrix, as needed later for DE
    rownames(df) <- df$cluster_sample_id
    
    ## Store complete metadata for cluster i in list
    metadata_ls[[i]] <- df
    names(metadata_ls)[i] <- unique(df$cluster_id)

}

# Explore the different components of the list
str(metadata_ls)
```
everything above was to prepare count matrices and sample-level metadata for each celltype

```{r}
# Select cell type of interest
cluster_names

# Double-check that both lists have same names
all(names(counts_ls) == names(metadata_ls))
```

```{r}
idx <- which(names(counts_ls) == "ISC/EB")
cluster_counts <- counts_ls[[idx]]
cluster_metadata <- metadata_ls[[idx]]
```

```{r}
# Check contents of extracted objects
cluster_counts
head(cluster_metadata)

# Check matching of matrix columns and metadata rows
all(colnames(cluster_counts) == rownames(cluster_metadata))
```

```{r}
dds <- DESeqDataSetFromMatrix(cluster_counts,
                              colData = cluster_metadata,
                              design = ~ group_id)
```

```{r}
# Transform counts for data visualization
rld <- rlog(dds, blind=TRUE)

# Plot PCA
DESeq2::plotPCA(rld, ntop = 500, intgroup = "group_id")
```

```{r}
DESeq2::plotPCA(rld, ntop = 500, intgroup = "cell_count")
```

```{r}
# Extract the rlog matrix from the object and compute pairwise correlation values
rld_mat <- assay(rld)
rld_cor <- cor(rld_mat)

# Plot heatmap
pheatmap(rld_cor, annotation = cluster_metadata[, c("group_id"), drop=F])
```

```{r}
# Run DESeq2 differential expression analysis
dds <- DESeq(dds)
```

```{r}
# Plot dispersion estimates
plotDispEsts(dds)
```

## Exploring DE results

```{r}
# Check the coefficients for the comparison
resultsNames(dds)

# Generate results object
res <- results(dds, 
               name = "group_id_rab11_vs_ctrl",
               alpha = 0.05)

# Shrink the log2 fold changes to be more appropriate using the apeglm method - should cite [paper]() when using this method
res <- lfcShrink(dds, 
                 coef = "group_id_rab11_vs_ctrl",
                 res=res,
                 type = "normal")
```

##Table of results for all genes

```{r}
# Turn the DESeq2 results object into a tibble for use with tidyverse functions
res_tbl <- res %>%
  data.frame() %>%
  rownames_to_column(var = "gene") %>%
  as_tibble() %>%
  arrange(padj)

# Check results output
res_tbl 
```


##Table of results for significant genes

```{r}
# Set thresholds
padj_cutoff <- 0.05

# Subset the significant results
sig_res <- dplyr::filter(res_tbl, padj < padj_cutoff) %>%
  dplyr::arrange(padj)

# Check significant genes output
sig_res
```

```{r}
# Set thresholds
log2fc_cutoff <- 0.58 #We may also be interested in determining the total number of significantly upregulated or downregulated genes above a certain fold change threshold (for example log2 fold change (in absolute value) >0.58, which corresponds to a ~50% increase (or ~30% decrease) in gene expression.

# Count significantly up/down genes above threshold
n_sig_up <- dplyr::filter(sig_res, log2FoldChange >= log2fc_cutoff) %>% 
  nrow()
sig_up <- dplyr::filter(sig_res, log2FoldChange >= log2fc_cutoff)
sig_up$gene
sig_dn <- dplyr::filter(sig_res, log2FoldChange <= log2fc_cutoff)
sig_dn$gene
n_sig_dn <- dplyr::filter(sig_res, log2FoldChange <= -log2fc_cutoff) %>% 
  nrow()
n_sig_up
n_sig_dn
```

##Scatterplot of normalized expression of top 20 most significant genes

```{r}
# Extract normalized counts from dds object
normalized_counts <- counts(dds, normalized = TRUE)

## Extract top 20 DEG from resLFC (make sure to order by padj)
top20_sig_genes <- sig_res %>%
  dplyr::arrange(padj) %>%
  dplyr::pull(gene) %>%
  head(n = 20)%>%
  c("rab11")
top20_sig_genes
```


```{r}
## Extract matching normalized count values from matrix
top20_sig_counts <- normalized_counts[rownames(normalized_counts) %in% top20_sig_genes, ]
top20_sig_counts
```


```{r}
## Convert wide matrix to long data frame for ggplot2
top20_sig_df <- data.frame(top20_sig_counts)
top20_sig_df$gene <- rownames(top20_sig_counts)

top20_sig_df <- melt(setDT(top20_sig_df), 
                     id.vars = c("gene"),
                     variable.name = "cluster_sample_id") %>% 
  data.frame()

## Replace "." by " " in cluster_sample_id variable (melt() introduced the ".")
top20_sig_df$cluster_sample_id <- gsub("\\.", "-", top20_sig_df$cluster_sample_id)
top20_sig_df
```


```{r}
## Join counts data frame with metadata
top20_sig_df <- plyr::join(top20_sig_df, as.data.frame(colData(dds)),
                           by = "cluster_sample_id")
top20_sig_df
```


```{r}
## Generate plot
top20_sig_df$group_id <- case_when(
  top20_sig_df$cluster_sample_id %in% c("DS-00178", "DS-00184") ~ "ctrl",
  top20_sig_df$cluster_sample_id %in% c("DS-0017A", "DS-00186") ~ "rab11",
)
top20_sig_df$group_id <- as.factor(top20_sig_df$group_id)
ggplot(top20_sig_df, aes(y = value, x = group_id, col = group_id)) +
  geom_jitter(height = 0, width = 0.15) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_y_continuous(trans = 'log10') +
  ylab("log10 of normalized expression level") +
  xlab("group_id") +
  ggtitle("Top 20 Significant DE Genes") +
  theme(plot.title = element_text(hjust = 0.5), aspect.ratio = 0.8) +
  facet_wrap(~ gene)
```
```{r}
# Adjust case_when to match the cluster_sample_id patterns correctly
top20_sig_df$group_id <- case_when(
  grepl("DS-00151", top20_sig_df$cluster_sample_id) ~ "ctrl",
  grepl("DS-00184", top20_sig_df$cluster_sample_id) ~ "ctrl",
  grepl("DS-00152", top20_sig_df$cluster_sample_id) ~ "rab11",
  grepl("DS-00187", top20_sig_df$cluster_sample_id) ~ "rab11",
  TRUE ~ NA_character_  # Handle cases where no match is found
)

# Convert group_id to factor
top20_sig_df$group_id <- as.factor(top20_sig_df$group_id)

# Inspect the structure and summary of the dataframe
str(top20_sig_df)
summary(top20_sig_df$group_id)

# Plot
library(ggplot2)
ggplot(top20_sig_df, aes(y = value, x = group_id, col = group_id)) +
  geom_jitter(height = 0, width = 0.15) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_y_continuous(trans = 'log10') +
  ylab("log10 of normalized expression level") +
  xlab("group_id") +
  ggtitle("Top 20 Significant DE Genes") +
  theme(plot.title = element_text(hjust = 0.5), aspect.ratio = 0.8) +
  facet_wrap(~ gene)

```

conclusion: 5 positive log2 fold changes for the top 10 significant genes(?) Negative log2 fold changes indicate that the gene is less expressed in the condition of interest compared to the control or reference condition.

```{r}
# Heatmap

## Extract normalized counts for significant genes only
sig_counts <- normalized_counts[rownames(normalized_counts) %in% sig_res$gene, ]

## Set a color-blind friendly palette
heat_colors <- rev(brewer.pal(11, "PuOr"))

## Run pheatmap using the metadata data frame for the annotation
pheatmap(sig_counts, 
         color = heat_colors, 
         cluster_rows = TRUE, 
         show_rownames = FALSE,
         annotation = cluster_metadata[, c("group_id", "cluster_id")], 
         border_color = NA, 
         fontsize = 10, 
         scale = "row", 
         fontsize_row = 10, 
         height = 20)  
```
conlusion: We see two clear modules of genes emerge, maybe we would get emphasizing the genes that are upregulated in the stimulated samples versus those that are downregulated if we had more samples

```{r}
# Volcano plot
res_table_thres <- res_tbl[!is.na(res_tbl$padj), ] %>% 
  mutate(threshold = padj < padj_cutoff & abs(log2FoldChange) >= log2fc_cutoff)
min(log10(res_table_thres$padj))

Top_Hits <- head(arrange(res_table_thres, pvalue), 20)

# Add column label, containing the gene name for the top hits or nothing for all others
res_table_thres$label <- if_else(res_table_thres$gene %in% Top_Hits$gene,  
                                 as.character(res_table_thres$gene), "")

## Generate plot
ggplot(res_table_thres) +
  geom_point(aes(x = log2FoldChange, y = -log10(padj), colour = threshold)) +
  ggtitle("Perturbed ISC relative to control") +
  xlab("log2 fold change") +
  xlim(-3, 3) +
  ylab("-log10 adjusted p-value") +
  scale_y_continuous(limits = c(0, 6)) +
  scale_color_manual(values = c("grey60", "red3")) +
  ggrepel::geom_text_repel(aes(x = log2FoldChange, y = -log10(padj), label = label),
                           size = 3, show.legend = FALSE,
                           max.overlaps = Inf)  +
  theme(legend.position = "none",
        plot.title = element_text(size = rel(1.3), hjust = 0.5),
        axis.title = element_text(size = rel(1.15)))
```

```{r, eval=FALSE}
#this is a mess ignore
# Identify the top 20 hits
Top_Hits <- head(arrange(res_table, pvalue), 20) #replaced res_table_thres with res_table
filtered_data <- res_table %>%
  filter(threshold == TRUE)
# Add column label, containing the gene name for the top hits or nothing for all others
filtered_data$label <- if_else(filtered_data$gene %in% Top_Hits$gene,  
                                 as.character(filtered_data$gene), "")

# Create the ggplot
ggplot(filtered_data, aes(x = log2FoldChange, y = -log10(padj), colour = threshold)) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = label),
                           size = 3, show.legend = FALSE,
                           max.overlaps = 200)  
true_threshold_values <- res_table_thres %>%
  filter(threshold == TRUE)

# Print or use the true_threshold_values as needed
print(true_threshold_values$gene)
```


Results interpretation: x-axis - log2 fold change for each gene (a measure of the magnitude of expression change between conditions) - more negative values that indicate downregulation (positive values indicate upregulation)
y-axis - the negative log10 of the p-value associated with the statistical test for each gene. This transformation helps visualize the statistical significance of gene expression changes. Higher values indicate lower p-values and, thus, greater statistical significance. we have the opposite :( 
gray - not sure

conclusions: we didn't find genes that are significantly differentially expressed (which are often located in the upper right and upper left quadrants)


20 sig genes
RpS19b
AstA-R1
CG43218
tinc
Cyp4p2
CG17778
LysD
mwh
CG11893
mgl
lz
CG31705
Cyp6a23
stl
CG4341
mfas
Mur18B
CG8620
Cyp6a9
psd

upreg genes
RpS19b
CG43218
tinc
Cyp4p2
lysD
mwh
CG11893
CG31705
stl

downreg genes
AstA-R1
CG17778
Mgl
Lz
Cyp6a23
CG4341
Mfas
Mur18B
CG8620
Cyp6a9
psd
