---
title: "annotation_flybase_ctrl"
output: html_document
date: "2023-11-02"
editor_options: 
  chunk_output_type: inline
---

# Introduction
This is some code based on what Nick Hirschmüller did to annotate his data with information from the fly cell atlas and to annotate manually using marker gene information. 
Some of this we will be able to

```{r, message=FALSE}
library(Seurat)
library(tidyverse)
library(SingleR)
library(patchwork)
library(circlize)
library(ComplexHeatmap)
library(qs)
set.seed(7)
```

```{r load data}
# loading the seurat object, for example from control

ds_00184_flybase <- readRDS(file = "/g/huber/users/lazoriak/ds_00184_control.rds")
ds_00185_flybase <- readRDS(file = "/g/huber/users/lazoriak/scripts/clustering_usp_new.rds")
ds_00186_flybase <- readRDS(file="/g/huber/users/lazoriak/scripts/clustering_dome.rds")
ds_00187_flybase <- readRDS(file="/g/huber/users/lazoriak/scripts/rab11/clustering_rab11.rds")

# loading the flycell_gut data:

flycell_gut <- qs::qread("/g/huber/users/hirschmueller/DECODE/raw_data/flycellatlas/gut_stringent/flycellatlas_gut_sce.qs")
```

```{r annotate, eval=FALSE}
# using singleR to do the annotation
# make sure this is accurate to your data and then you will have to run this on the cluster / not in Rstudio

labels <- SingleR::SingleR(test=ds_00184@assays$RNA@counts,
                  ref =SingleCellExperiment::counts(flycell_gut),
                  labels = flycell_gut$annotation,
                  de.method="wilcox",
                  BPPARAM=MulticoreParam(16))
```

```{r}
# load the annotaiton after you're done on the cluster
# add the labels to your seurat object similarly as you did before

labels <- readRDS(file = "/g/huber/users/lazoriak/scripts/integration/annotation_results/labels.rds")
labels_usp <- readRDS(file="/g/huber/users/lazoriak/scripts/integration/annotation_results/labels_usp.rds")
labels_dome <- readRDS(file="/g/huber/users/lazoriak/scripts/integration/annotation_results/labels_dome.rds")
labels_rab11 <- readRDS(file="/g/huber/users/lazoriak/scripts/integration/annotation_results/labels_rab11.rds")
```

```{r}
# use new names, as to not overwrite the other annotations, so we can compare them
# maybe use ds_00184@meta.data$flyatlas_label & use labels$labels and labels$pruned.labels

ds_00184_flybase@meta.data$flyatlas_label <- labels$labels
ds_00184_flybase@meta.data$flyatlas_label_pruned <- labels$pruned.labels

ds_00185_flybase@meta.data$flyatlas_label <- labels_usp$labels
ds_00185_flybase@meta.data$flyatlas_label_pruned <- labels_usp$pruned.labels

ds_00186_flybase@meta.data$flyatlas_label <- labels_dome$labels
ds_00186_flybase@meta.data$flyatlas_label_pruned <- labels_dome$pruned.labels

ds_00187_flybase@meta.data$flyatlas_label <- labels_rab11$labels
ds_00187_flybase@meta.data$flyatlas_label_pruned <- labels_rab11$pruned.labels
```

```{r}
DimPlot(ds_00184_flybase, reduction = "umap", group.by="flyatlas_label", label=FALSE) +
  theme(legend.text = element_text(size = 8))
```

# Integrating usp perturbation annotation results with the control data

```{r}
# making a list of Seurat objects
samples_list <- list(ds_00184_flybase, ds_00185_flybase)
# normalizing and identifying variable features for each dataset independently
samples_list <- lapply(X = samples_list, FUN = function(x) {
    x <- NormalizeData(x)
    x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 2000)
})

# selecting features that are repeatedly variable across datasets for integration
features <- SelectIntegrationFeatures(object.list = samples_list)

# integration
anchors_new <- FindIntegrationAnchors(object.list = samples_list, anchor.features = features)

# this command creates an 'integrated' data assay
usp_assay_combined <- IntegrateData(anchorset = anchors_new)
DefaultAssay(usp_assay_combined) <- "integrated"

usp_assay_combined <- ScaleData(usp_assay_combined, verbose = FALSE)
usp_assay_combined <- RunPCA(usp_assay_combined, npcs = 30, verbose = FALSE)
usp_assay_combined <- RunUMAP(usp_assay_combined, reduction = "pca", dims = 1:30)
usp_assay_combined <- FindNeighbors(usp_assay_combined, reduction = "pca", dims = 1:30)
usp_assay_combined <- FindClusters(usp_assay_combined, resolution = 0.5)
```

```{r}
# visualize these annotations and compare to previous annotations... how are they different?
# here it would be useful to plot multiple UMAPs at once, one with each annotation
# we can do this together also
p1 <- DimPlot(usp_assay_combined, reduction = "umap", group.by = "orig.ident") #ident or?
p2 <- DimPlot(usp_assay_combined, reduction = "umap", label = TRUE, repel = TRUE)
p1
p2
```

```{r}
DimPlot(usp_assay_combined, reduction = "umap", split.by = "orig.ident")
```
```{r}

# the column "labels_combined" now has all the label information from labels and labels.x
# you should now be able to use that column to color your DimPlot by cell type, probably with the parameter "group.by="
DimPlot(usp_assay_combined, 
        reduction = "umap", 
        split.by = "orig.ident", 
        group.by = "flyatlas_label") +
  theme(legend.text = element_text(size = 8))
```
# Integrating dome perturbation annotation results with the control data

```{r}
# making a list of Seurat objects
samples_list <- list(ds_00184_flybase, ds_00186_flybase)
# normalizing and identifying variable features for each dataset independently
samples_list <- lapply(X = samples_list, FUN = function(x) {
    x <- NormalizeData(x)
    x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 2000)
})

# selecting features that are repeatedly variable across datasets for integration
features <- SelectIntegrationFeatures(object.list = samples_list)

# integration
anchors_new <- FindIntegrationAnchors(object.list = samples_list, anchor.features = features)

# this command creates an 'integrated' data assay
dome_assay_combined <- IntegrateData(anchorset = anchors_new)
DefaultAssay(dome_assay_combined) <- "integrated"

dome_assay_combined <- ScaleData(dome_assay_combined, verbose = FALSE)
dome_assay_combined <- RunPCA(dome_assay_combined, npcs = 30, verbose = FALSE)
dome_assay_combined <- RunUMAP(dome_assay_combined, reduction = "pca", dims = 1:30)
dome_assay_combined <- FindNeighbors(dome_assay_combined, reduction = "pca", dims = 1:30)
dome_assay_combined <- FindClusters(dome_assay_combined, resolution = 0.5)
```

```{r}
# visualize these annotations and compare to previous annotations... how are they different?
# here it would be useful to plot multiple UMAPs at once, one with each annotation
# we can do this together also
p1 <- DimPlot(dome_assay_combined, reduction = "umap", group.by = "orig.ident") #ident or?
p2 <- DimPlot(dome_assay_combined, reduction = "umap", label = TRUE, repel = TRUE)
p1
p2
```

```{r}
DimPlot(dome_assay_combined, reduction = "umap", split.by = "orig.ident")
```
```{r}
DimPlot(dome_assay_combined, reduction = "umap", split.by = "orig.ident", group.by="flyatlas_label")+
  theme(legend.text = element_text(size = 8))
```

# Integrating rab11 perturbation annotation results with the control data

```{r}
# making a list of Seurat objects
samples_list <- list(ds_00184_flybase, ds_00187_flybase)
# normalizing and identifying variable features for each dataset independently
samples_list <- lapply(X = samples_list, FUN = function(x) {
    x <- NormalizeData(x)
    x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 2000)
})

# selecting features that are repeatedly variable across datasets for integration
features <- SelectIntegrationFeatures(object.list = samples_list)

# integration
anchors_new <- FindIntegrationAnchors(object.list = samples_list, anchor.features = features)

# this command creates an 'integrated' data assay
rab11_assay_combined <- IntegrateData(anchorset = anchors_new)
DefaultAssay(rab11_assay_combined) <- "integrated"

rab11_assay_combined <- ScaleData(rab11_assay_combined, verbose = FALSE)
rab11_assay_combined <- RunPCA(rab11_assay_combined, npcs = 30, verbose = FALSE)
rab11_assay_combined <- RunUMAP(rab11_assay_combined, reduction = "pca", dims = 1:30)
rab11_assay_combined <- FindNeighbors(rab11_assay_combined, reduction = "pca", dims = 1:30)
rab11_assay_combined <- FindClusters(rab11_assay_combined, resolution = 0.5)
```


```{r}
# visualize these annotations and compare to previous annotations... how are they different?
# here it would be useful to plot multiple UMAPs at once, one with each annotation
# we can do this together also
p1 <- DimPlot(rab11_assay_combined, reduction = "umap", group.by = "orig.ident") #ident or?
p2 <- DimPlot(rab11_assay_combined, reduction = "umap", label = TRUE, repel = TRUE)
p1
p2
```

```{r}
DimPlot(rab11_assay_combined, reduction = "umap", split.by = "orig.ident")
```

```{r}
DimPlot(rab11_assay_combined, reduction = "umap", split.by = "orig.ident", group.by="flyatlas_label")+
  theme(legend.text = element_text(size = 8))
```

```{r}
# save annotated data
saveRDS (usp_assay_combined, file="/g/huber/users/lazoriak/scripts/integration/annotation_results/ctrl_usp_annotation_combined.rds")
saveRDS (dome_assay_combined, file="/g/huber/users/lazoriak/scripts/integration/annotation_results/ctrl_dome_annotation_combined.rds")
saveRDS (rab11_assay_combined, file="/g/huber/users/lazoriak/scripts/integration/annotation_results/ctrl_rab11_annotation_combined.rds")
```