---
title: "integrated_perturbated_samples_and_control"
output: html_document
date: "2023-11-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#downloading all libraries and data we will need :)
library(Seurat)
library(SeuratData) #to install: devtools::install_github("satijalab/seurat-data", ref = 'develop')
library(patchwork)

#control sample
ds_00184.data <- Read10X(data.dir = "/g/huber/users/philipp/decode/data/DS-00184/counts/outs/filtered_feature_bc_matrix/")
ds_00184 <- CreateSeuratObject(counts = ds_00184.data, project = "decode_ctrl", min.cells = 3, min.features = 200)

#usp
ds_00185.data <- Read10X(data.dir = "/g/huber/users/philipp/decode/data/DS-00185/counts/outs/filtered_feature_bc_matrix/")
ds_00185 <- CreateSeuratObject(counts = ds_00185.data, project = "decode_usp", min.cells = 3, min.features = 200)

#dome
ds_00186.data <- Read10X(data.dir = "/g/huber/users/philipp/decode/data/DS-00186/counts/outs/filtered_feature_bc_matrix/")
ds_00186 <- CreateSeuratObject(counts = ds_00186.data, project = "decode_dome", min.cells = 3, min.features = 200)

#rab11
ds_00187.data <- Read10X(data.dir = "/g/huber/users/philipp/decode/data/DS-00187/counts/outs/filtered_feature_bc_matrix/")
ds_00187 <- CreateSeuratObject(counts = ds_00187.data, project = "decode_rab11", min.cells = 3, min.features = 200)
```

# Integrating usp perturbation samples with the control data

## Setting up the Seurat objects

```{r}
# making a list of Seurat objects for downstream analysis
seurat_object_list <- list(ds_00184, ds_00185)

# normalizing and identifying variable features for each dataset independently
seurat_object_list <- lapply(X = seurat_object_list, FUN = function(x) {
    x <- NormalizeData(x)
    x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 2000)
})

# selecting features that are repeatedly variable across datasets for integration
features <- SelectIntegrationFeatures(object.list = seurat_object_list)
```

## Performing integration

```{r}
# integration
anchors <- FindIntegrationAnchors(object.list = seurat_object_list, anchor.features = features)

# this command creates an 'integrated' data assay (why?)
ctrl_usp_assay_combined <- IntegrateData(anchorset = anchors)
```

## Performing an integrated analysis

```{r}
# specifying that we will perform downstream analysis on the corrected data, note that the
# original unmodified data still resides in the 'RNA' assay
DefaultAssay(ctrl_usp_assay_combined) <- "integrated"

# Running the standard workflow for visualization and clustering
ctrl_usp_assay_combined <- ScaleData(ctrl_usp_assay_combined, verbose = FALSE)
ctrl_usp_assay_combined <- RunPCA(ctrl_usp_assay_combined, npcs = 30, verbose = FALSE)
ctrl_usp_assay_combined <- RunUMAP(ctrl_usp_assay_combined, reduction = "pca", dims = 1:30)
ctrl_usp_assay_combined <- FindNeighbors(ctrl_usp_assay_combined, reduction = "pca", dims = 1:30)
ctrl_usp_assay_combined <- FindClusters(ctrl_usp_assay_combined, resolution = 0.5)

# Visualization
p1 <- DimPlot(ctrl_usp_assay_combined, reduction = "umap", group.by = "orig.ident") #ident or?
p2 <- DimPlot(ctrl_usp_assay_combined, reduction = "umap", label = TRUE, repel = TRUE)
p1 + p2
```
```{r}
DimPlot(ctrl_usp_assay_combined, reduction = "umap", label=TRUE, split.by = "orig.ident")
# To visualize the two conditions side-by-side, we can use the split.by argument to show each condition colored by cluster. (why do we need it?)
```
```{r}
#saving the integrated data
saveRDS (ctrl_usp_assay_combined, file="/g/huber/users/lazoriak/scripts/integration/ctrl_usp_assay_combined.rds")
```

# Integrating dome perturbation samples with the control data

## Setting up the Seurat objects

```{r}
# making a list of Seurat objects for downstream analysis
seurat_object_list_2 <- list(ds_00184, ds_00186)

# normalizing and identifying variable features for each dataset independently
seurat_object_list_2 <- lapply(X = seurat_object_list_2, FUN = function(x) {
    x <- NormalizeData(x)
    x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 2000)
})

# selecting features that are repeatedly variable across datasets for integration
features <- SelectIntegrationFeatures(object.list = seurat_object_list_2)
```

## Performing integration

```{r}
# integration
anchors <- FindIntegrationAnchors(object.list = seurat_object_list_2, anchor.features = features)

# this command creates an 'integrated' data assay (why?)
ctrl_dome_assay_combined <- IntegrateData(anchorset = anchors)
```

## Performing an integrated analysis

```{r}
# specifying that we will perform downstream analysis on the corrected data, note that the
# original unmodified data still resides in the 'RNA' assay
DefaultAssay(ctrl_dome_assay_combined) <- "integrated"

# Running the standard workflow for visualization and clustering
ctrl_dome_assay_combined <- ScaleData(ctrl_dome_assay_combined, verbose = FALSE)
ctrl_dome_assay_combined <- RunPCA(ctrl_dome_assay_combined, npcs = 30, verbose = FALSE)
ctrl_dome_assay_combined <- RunUMAP(ctrl_dome_assay_combined, reduction = "pca", dims = 1:30)
ctrl_dome_assay_combined <- FindNeighbors(ctrl_dome_assay_combined, reduction = "pca", dims = 1:30)
ctrl_dome_assay_combined <- FindClusters(ctrl_dome_assay_combined, resolution = 0.5)

# Visualization
p1 <- DimPlot(ctrl_dome_assay_combined, reduction = "umap", group.by = "orig.ident") #ident or?
p2 <- DimPlot(ctrl_dome_assay_combined, reduction = "umap", label = TRUE, repel = TRUE)
p1 
p2
```
```{r}
DimPlot(ctrl_dome_assay_combined, reduction = "umap", split.by = "orig.ident")
# To visualize the two conditions side-by-side, we can use the split.by argument to show each condition colored by cluster. (why do we need it?)
```
```{r}
#saving the integrated data
saveRDS (ctrl_dome_assay_combined, file="/g/huber/users/lazoriak/scripts/integration/ctrl_dome_assay_combined.rds")
```

# Integrating rab11 perturbation samples with the control data

## Setting up the Seurat objects

```{r}
# making a list of Seurat objects for downstream analysis
seurat_object_list_3 <- list(ds_00184, ds_00187)

# normalizing and identifying variable features for each dataset independently
seurat_object_list_3 <- lapply(X = seurat_object_list_3, FUN = function(x) {
    x <- NormalizeData(x)
    x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 2000)
})

# selecting features that are repeatedly variable across datasets for integration
features <- SelectIntegrationFeatures(object.list = seurat_object_list_3)
```

## Performing integration

```{r}
# integration
anchors <- FindIntegrationAnchors(object.list = seurat_object_list_3, anchor.features = features)

# this command creates an 'integrated' data assay (why?)
ctrl_rab11_assay_combined <- IntegrateData(anchorset = anchors)
```

## Performing an integrated analysis

```{r}
# specifying that we will perform downstream analysis on the corrected data, note that the
# original unmodified data still resides in the 'RNA' assay
DefaultAssay(ctrl_rab11_assay_combined) <- "integrated"

# Running the standard workflow for visualization and clustering
ctrl_rab11_assay_combined <- ScaleData(ctrl_rab11_assay_combined, verbose = FALSE)
ctrl_rab11_assay_combined <- RunPCA(ctrl_rab11_assay_combined, npcs = 30, verbose = FALSE)
ctrl_rab11_assay_combined <- RunUMAP(ctrl_rab11_assay_combined, reduction = "pca", dims = 1:30)
ctrl_rab11_assay_combined <- FindNeighbors(ctrl_rab11_assay_combined, reduction = "pca", dims = 1:30)
ctrl_rab11_assay_combined <- FindClusters(ctrl_rab11_assay_combined, resolution = 0.5)

# Visualization
p1 <- DimPlot(ctrl_rab11_assay_combined, reduction = "umap", group.by = "orig.ident") #ident or?
p2 <- DimPlot(ctrl_rab11_assay_combined, reduction = "umap", label = TRUE, repel = TRUE)
p1 + p2
```
```{r}
DimPlot(ctrl_rab11_assay_combined, reduction = "umap", split.by = "orig.ident")
# To visualize the two conditions side-by-side, we can use the split.by argument to show each condition colored by cluster. (why do we need it?)
```
```{r}
#saving the integrated data
saveRDS (ctrl_rab11_assay_combined, file="/g/huber/users/lazoriak/scripts/integration/ctrl_rab11_assay_combined.rds")
```