---
title: "trajectory inference"
output: html_document
date: "2024-03-12"
---


## Trajectory inference

* what is it and why? -> maybe this paper is helpful: https://www.nature.com/articles/s41467-020-14766-3
* many different packages for this, we start with slingshot
* slingshot paper: https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-018-4772-0
* slingshot tutorial / manual: https://bioconductor.org/packages/devel/bioc/vignettes/slingshot/inst/doc/vignette.html 
* condiments is slingshot for multiple conditions
* condiments paper: https://www.biorxiv.org/content/10.1101/2021.03.09.433671v1.full
* condiments tutorial / manual: https://www.bioconductor.org/packages/release/bioc/vignettes/condiments/inst/doc/condiments.html

* tradeSeq is for differentially expressed genes in trajectory analysis (there is info on tradeSeq in the slingshot manual)

* this would be conducted on data annotated with cell types

```{r}
library(condiments)
```

```{r}
slingshot_res_nick <- readRDS("/g/huber/users/hirschmueller/DECODE/NotchPaper/code/scNotch_profiling/output/slingshot_object.rds")
top_nick <- topologyTest(SlingshotDataSet(slingshot_res_nick), 
                                    slingshot_res_nick$perturbation, 
                                    rep = 50,
                                    methods = "KS_mean", 
                                    threshs = .01,
                                    parallel = TRUE,
                                    BPPARAM = BiocParallel::MulticoreParam(workers = 50))
```

```{r setup, include=FALSE}
 # load your annotated integrated rds here
ctrl_combined <- merge(ctrl_rep1$srt, y=ctrl_rep2$srt, add.cell.ids = c("ctrl_1", "ctrl_2"))

# combine "treatment"
usp_combined <- merge(usp_rep1$srt, y=usp_rep2$srt, add.cell.ids = c("usp_1","usp_2"))

# combine all
usp_ctrl <- merge(ctrl_combined, y=usp_combined)
usp_ctrl@meta.data$sample_id <- factor(usp_ctrl@meta.data$orig.ident)
usp_ctrl@meta.data$group_id <- case_when(
  usp_ctrl@meta.data$sample_id %in% c("DS-00151", "DS-00184") ~ "ctrl",
  usp_ctrl@meta.data$sample_id %in% c("DS-00157", "DS-00185") ~ "usp",
                                      ) 
usp_ctrl$group_id <- factor(usp_ctrl$group_id)
saveRDS(usp_ctrl, file="/g/huber/users/lazoriak/scripts/trajectories/usp_ctrl.rds")
```

```{r analysis before slingshot}
# slingshot requires filtering, normalization, dimensionality reduction and clustering
# which we have all done before

# MAYBE: reducing the number of cell types (likely already done)
# MAYBE: reintegration if number of cell types changed
```

```{r convert to sce}
# convert seurat object to sce
sce <- as.SingleCellExperiment(usp_ctrl)
```

```{r}
sce <- scater::logNormCounts(sce)
hvg <- order(MatrixGenerics::rowVars(logcounts(sce)), decreasing = TRUE)
sce <- sce[hvg[1:500], ]
sce <- scater::runPCA(sce, ncomponents = 50)
sce <- scater::runUMAP(sce, dimred = "PCA")
saveRDS(sce, file="/g/huber/users/lazoriak/scripts/trajectories/sce_before_slingshot.rds")
```

```{r use slingshot}
# might have to define starting and ending cell types (starting celltype would be the stem cells)

sce <- slingshot(sce, clusterLabels = 'predicted.celltype', reducedDim = 'PCA')

# this will be fitting one trajectory for both ctrl and perturbation

saveRDS(sce, file="/g/huber/users/lazoriak/scripts/trajectories/sce_after_slingshot.rds")
```

```{r visualise outcome}
# if need be, reload slingshot results
#sce <- readRDS(file="/g/huber/users/lazoriak/scripts/trajectories/sce_after_slingshot.rds")
# visualise the outcome
summary(sce$slingPseudotime_1)
library(raster)
library(rgdal)
library(classInt)
library(RColorBrewer)
library(grDevices)
```

```{r visualise outcome}
colors <- colorRampPalette(brewer.pal(11,'Spectral')[-6])(100)
plotcol <- colors[cut(sce$slingPseudotime_1, breaks=100)]
plot(reducedDims(sce)$PCA, col = plotcol, pch=16, asp = 1)
lines(SlingshotDataSet(sce), lwd=2, col='black')
```

```{r visualise outcome}
plot(reducedDims(sce)$PCA, col = brewer.pal(9,'Set1')[sce$GMM], pch=16, asp = 1)
lines(SlingshotDataSet(sce), lwd=2, type = 'lineages', col = 'black')
```



```{r}
# we might have to adjust the fit or fit two different trajectories
```

```{r}
# now we can test for differentially expressed genes throughout the trajectories
# for that we need to use tradeSeq
library(tradeSeq)

# fit negative binomial GAM
sce <- fitGAM(sce)

# test for dynamic expression
ATres <- associationTest(sce)
```

```{r}
saveRDS(sce, file="/g/huber/users/lazoriak/scripts/trajectories/sce_after_fitGAM.rds")
topgenes <- rownames(ATres[order(ATres$pvalue), ])[1:250]
pst.ord <- order(sce$slingPseudotime_1, na.last = NA)
heatdata <- assays(sce)$counts[topgenes, pst.ord]
heatclus <- sce$predicted.celltype[pst.ord]

heatmap(as.matrix(log1p(heatdata)), Colv = NA,
        ColSideColors = brewer.pal(9,"Set1")[heatclus])
```

```{r}
# take a look at the genes that come out of this analysis
# can be tested with enrichr
# or compared to the previous analyses
```

```{r}
# Authors of condiments/slingshot prefer fitting a common trajectory if the differences
# between the conditions is small enough!
# here we check if this is the case for our data
slingshot_res <- readRDS(file="/g/huber/users/lazoriak/scripts/trajectories/sce_after_slingshot.rds")

scores <- condiments::imbalance_score(
  Object = reducedDim(sce,"PCA")[,1:2] %>% as.matrix(),
  conditions = sce$group_id,
  k = 40, smooth = 30)

sce$scores <- scores$scaled_scores
imbalence_df<-data.frame(reducedDim(sce,"PCA")[,1:2])
imbalence_df$scores <- scores$scaled_scores

# we can see, that there are areas of great imbalence. Which is expected given we see different celltype abundance.
ggplot(imbalence_df, aes(x = PC1, y = PC2, col = scores)) +
  geom_point(size = .7) +
  scale_color_viridis_c(option = "C") +
  labs(col = "Scores")

# essentially, the labels get permutated and the trajectories are refit.
# pseudotime vector is generated and the distribution is compared with the "true" one
# if the results do not change significantly, we can fit the same trajectory for both conditions.

# This result suggests we should fit seperat trajetories for the conditions
topologyTest_result <- topologyTest(PseudotimeOrdering(slingshot_res), 
                                    slingshot_res$group_id, 
                                    rep = 120,
                                    methods = "KS_mean",
                                    threshs = .01,
                                    parallel = TRUE,
                                    BPPARAM = BiocParallel::MulticoreParam(workers = 50))
topologyTest_result <- readRDS(file="/g/huber/users/lazoriak/scripts/trajectories/topologyTest_result.rds")

# refit slingshot
slingshot_individual <- slingshot_conditions(slingshot_res,
                                             slingshot_res$group_id,
                                             approx_points = 150)
```


